Polka-dotted Stars: a Hierarchical Model for Mapping Stellar Surfaces Using Occultation Light Curves and the Case of TOI-3884

### 2025-02-28 src/tex/ms.tex
wrote up discussion

@@

%
\begin{figure*}[hbt!]
    \centering
    \includegraphics[width=\textwidth]{figures/toi3884-samples-traj.pdf}
    \caption{Observer's view of TOI-3884 during transit events from nine representative posterior samples. 
    Each panel shows the stellar disk as viewed from Earth, with dark purple regions indicating cooler spots and 
    yellow-orange areas representing the warmer photosphere. Black dots trace the path of TOI-3884b across the stellar disk 
    during transit. The consistent presence of high-latitude spots intersecting or near the planet's trajectory explains the 
    asymmetries observed in the transit light curves. This visualization demonstrates how the combination of stellar inclination 
    ($i_\star = {34.8}^{+5.62}_{-6.17}$ degrees), spot distribution, and planetary orbital geometry produces the distinctive
    transit signatures in the TESS data.}
    \label{fig:toi3884-traj}
\end{figure*}
%
\bibliography{bib}

\end{document}

>>>>

\author[0000-0003-2528-3409]{Brett M. Morris}
% \email{wfarr@flatironinstitute.org}
\affiliation{Space Telescope Science Institute, 3700 San Martin Dr, Baltimore, MD 21218, USA}

@@

%
\begin{figure*}[hbt!]
    \centering
    \includegraphics[width=\textwidth]{figures/toi3884-samples-traj.pdf}
    \caption{Observer's view of TOI-3884 during transit events from nine representative posterior samples. 
    Each panel shows the stellar disk as viewed from Earth, with dark purple regions indicating cooler spots and 
    yellow-orange areas representing the warmer photosphere. Black dots trace the path of TOI-3884b across the stellar disk 
    during transit. The consistent presence of high-latitude spots intersecting or near the planet's trajectory explains the 
    asymmetries observed in the transit light curves. This visualization demonstrates how the combination of stellar inclination 
    ($i_\star = {34.8}^{+5.62}_{-6.17}$ degrees), spot distribution, and planetary orbital geometry produces the distinctive
    transit signatures in the TESS data.}
    \label{fig:toi3884-traj}
\end{figure*}
%

\section{Discussion}

Our analysis demonstrates that transit light curves can provide meaningful constraints on stellar surface features and 
three-dimensional geometry, even when using single-band photometric data. The model successfully recovers a consistent spot 
distribution across multiple posterior samples (see Section \ref{sec:experiment1}), characterized predominantly by high-latitude 
features in the case of TOI-3884 (see Section \ref{sec:toi3884}). 
This result is robust across our MCMC exploration and explains the observed transit light curve morphologies. 
However, there are several important considerations regarding the model's limitations and strengths.

\subsection{Parameter Degeneracies and Uncertainty}

While our model accurately reproduces the observed transit light curves, the posterior distributions for several key 
parameters remain relatively wide. This breadth reflects inherent degeneracies in transit modeling that persist even 
with our approach described in Section \ref{sec:model}. For instance, the stellar inclination angle 
($i_\star$) and sky-projected spin-orbit angle 
($\lambda_\star$) have substantial uncertainties. These uncertainties stem 
from multiple parameter combinations that can produce similar transit signatures, particularly regarding the precise spot 
locations and their contrast with the surrounding photosphere.

The degeneracy between northern and southern hemisphere spot locations is particularly evident. As shown in Figure 
\ref{fig:spot_latitude} and also discussed in Section \ref{sec:model}, we cannot definitively distinguish 
between spots located at positive or negative latitudes using 
transit data alone. This ambiguity is fundamental to the transit method, as the planet's path across the stellar disk 
cannot discriminate between spots above or below its trajectory when they are symmetric about the transit chord.

Despite these degeneracies, our flexible spot model recovers a consistent overall picture of TOI-3884's surface features. 
The near-polar spot configuration emerges as a robust result regardless of which specific parameter combination within our 
posterior is considered. This consistency suggests that while individual parameter values may be uncertain, the general 
surface feature distribution is well-constrained by the data.

\subsection{Recovery of 3D Information from Photometry Alone}

A significant achievement of our approach is the recovery of three-dimensional information about the star-planet system using 
only single-band photometry. Previous studies typically require complementary data sources such as Doppler tomography or 
spectropolarimetric observations to constrain the three-dimensional geometry of transiting systems. 
Our work demonstrates that with appropriate modeling, transit light curves themselves contain significant information about 
the stellar rotation axis and surface feature distribution.

Our derived stellar projected rotational velocity ($v\sin i_\star = {5.2}^{+0.7}_{-0.8}$ km/s) is in good agreement 
with spectroscopically measured values for TOI-3884 in \cite{Libby-Roberts2023}, despite being derived s
olely from photometric data. This agreement provides strong independent validation of our geometric model 
and suggests that photometric analyses can serve as useful complements to spectroscopic methods for determining stellar 
rotational properties.

Our results highlight important limitations in previous spot modeling approaches. For instance, 
the models employed before assumed either a single spot or a fixed number of discrete spots 
\citep{Morris2018,Oshagh2013,Libby-Roberts2023,Beky2014}. While computationally efficient, 
such strict assumptions artificially reduce uncertainty about the stellar orientation and spot configuration. 
By imposing strong priors on the number and morphology of spots, these models may underestimate the true parameter 
uncertainties and potentially miss important surface features.
Our Gaussian process-based approach allows for more flexible spot distributions with varying complexity, 
providing a more realistic assessment of the constraints that can be derived from transit data. 
The wider posterior distributions in our analysis more accurately reflect the genuine uncertainties inherent in 
inferring stellar surface features from transit light curves alone.

Figure \ref{fig:toi3884-compare} presents the posterior distributions and correlations for selected parameters in our model of 
TOI-3884. The parameters shown include the planet's inclination ($i_p$), orbital eccentricity ($e$), 
stellar inclination ($i_\star$), sky-projected spin-orbit angle ($\lambda_\star$), planet-to-star radius ratio ($R_p/R_\star$), 
projected stellar rotational velocity ($v\sin i$), and stellar density ($\rho_\star$).
The red shaded regions and vertical dashed lines indicate independent constraints from radial velocity measurements reported by 
\cite{Libby-Roberts2023}. These RV-derived constraints provide important independent validation for our photometry-based model. 
Our photometrically derived $v\sin i$ value of approximately 2.5 km/s aligns within 1$\sigma$-well with the spectroscopic 
measurement, 
despite being derived solely from transit light curves. This agreement strongly reinforces the power of our approach to 
extract detailed stellar and orbital parameters using only photometric data.
We show that with single-band photometry alone, we can constrain parameters traditionally measured through spectroscopic techniques. 
The ability to recover $v\sin i$ and place meaningful constraints on the spin-orbit alignment without recourse to 
spectroscopy represents a significant methodological advancement, potentially enabling detailed characterization of 
systems for which high-resolution spectroscopy is challenging.

\begin{figure*}[hbt!]
    \centering
    \includegraphics[width=\textwidth]{figures/toi3884-libby-compare.pdf}
    \caption{The corner plot for planet's inclination ($i_p$), orbital eccentricity ($e$), 
    stellar inclination ($i_\star$), sky-projected spin-orbit angle ($\lambda_\star$), planet-to-star radius ratio ($R_p/R_\star$), 
    projected stellar rotational velocity ($v\sin i$), and stellar density ($\rho_\star$). The red shaded regions and vertical dashed lines indicate independent constraints from radial velocity measurements reported by 
    \cite{Libby-Roberts2023}.}
    \label{fig:toi3884-compare}
\end{figure*}
%
\subsection{Implications for M Dwarf Magnetic Activity}

The high-latitude spot configuration we detect aligns with theoretical expectations for rapidly rotating low-mass stars like 
TOI-3884. Magnetohydrodynamic simulations predict that Coriolis forces in such stars should preferentially channel 
magnetic flux emergence toward the rotational poles \citep{Yadav2015,Almenara2022}. Our results provide observational support for these theoretical predictions, 
demonstrating that transit modeling can be a valuable tool for studying stellar magnetic activity patterns.
The combination of a moderately inclined stellar rotation axis and significant high-latitude spots also helps explain why 
previous analyses of TOI-3884 detected relatively low photometric variability despite evidence for active chromospheric emission. 
High-latitude spots maintain a relatively constant projected area as the star rotates when viewed at moderate inclinations, 
resulting in reduced rotational modulation while still affecting transit profiles when intersected by the planet's path.

\subsection{Future Directions}

While our analysis demonstrates the power of photometric transit modeling, the remaining degeneracies point 
to the value of multi-wavelength observations for future studies. Transit observations at different wavelengths would 
provide additional constraints on spot temperatures and coverage, potentially breaking some of the degeneracies in our current 
model. Similarly, combining our photometric approach with spectroscopic measurements would enable more precise determinations 
of the three-dimensional system geometry.
For TOI-3884 specifically, continued monitoring of transit events could reveal temporal evolution of the spot distribution, 
providing insights into the magnetic activity cycle of this M dwarf. Such observations would be valuable for understanding 
how stellar activity might impact the planetary environment and potential habitability of planets orbiting active M dwarfs.

\bibliography{bib}
\end{document}
